---
title: "FPR versus SSP comparison for screening statistics"
author: "mfscreen2 simulation"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 4.8
)

library(ggplot2)

result_dir <- "fpr_ssp_n2000_results"
summary_results <- read.csv(file.path(result_dir, "summary_fpr_ssp.csv"))
summary_results$q_label <- ifelse(
  summary_results$q %in% c(0.001, 0.01, 0.05, 0.10, 0.30),
  as.character(summary_results$q),
  ""
)

stress_dir <- "fpr_control_common_volatility_n2000_results"
stress_results <- read.csv(file.path(stress_dir, "summary_fpr_control.csv"))

q_levels <- sort(unique(summary_results$q))
n <- 2000L
p <- 1000L
s <- 5L
n_rep <- length(unique(read.csv(file.path(result_dir, "raw_fpr_ssp.csv"))$replication))
stress_n_rep <- length(unique(read.csv(file.path(stress_dir, "raw_fpr_control.csv"))$replication))
```

# Competing Screening Statistics

The simulation compares three marginal screening procedures. Each procedure
assigns one score to each predictor variable, then selects variables according
to a threshold indexed by a nominal level \(q\).

## Reciprocal model-free screening statistic

For predictor \(X_j\), let
\[
z_{ij} = X_{ij} - \bar X_j, \qquad u_i = Y_i - \bar Y.
\]
The `mfscreen` statistic is
\[
D_j^2 =
\frac{\sum_i z_{ij}^2 u_i^2}{\left(\sum_i z_{ij}u_i\right)^2}
-
\frac{2\sum_i z_{ij}^3 u_i}
{\left(\sum_i z_{ij}u_i\right)\left(\sum_i z_{ij}^2\right)}
+
\frac{\sum_i z_{ij}^4}{\left(\sum_i z_{ij}^2\right)^2}.
\]
It is the squared reciprocal of the robust marginal screening statistic:
\[
D_j^2 = \frac{1}{T_j^2}.
\]
Small values indicate stronger marginal evidence. For each nominal level \(q\),
we set
\[
\gamma_q = \Phi^{-1}(1 - q/2)
\]
and select variable \(j\) when
\[
D_j^2 \leq \frac{1}{\gamma_q^2}.
\]

## Pearson SIS

Pearson SIS is the classical marginal correlation screen. For each predictor,
we compute the Pearson correlation \(r_j\) between \(X_j\) and \(Y\), then use
the usual two-sided correlation test statistic
\[
t_j = |r_j|\sqrt{\frac{n-2}{1-r_j^2}}.
\]
The two-sided p-value is
\[
p_j = 2P(t_{n-2} \geq t_j).
\]
For nominal level \(q\), Pearson SIS selects predictor \(j\) when
\[
p_j \leq q.
\]

## Spearman SIS

Spearman SIS applies the same p-value rule after replacing each variable by
its ranks. It is a rank-based marginal screen and is less sensitive to monotone
transformations than Pearson correlation. For nominal level \(q\), it selects
predictor \(j\) when the two-sided rank-correlation p-value satisfies
\[
p_j^{\mathrm{rank}} \leq q.
\]

## How the plotted points are constructed

For a fixed method and a fixed nominal level \(q\), each Monte Carlo
replication produces a selected set \(\widehat S_q\). Let \(S\) be the true
active set and \(S^c\) its complement. In replication \(b\),
\[
\mathrm{FPR}_b(q) =
\frac{|\widehat S_q^{(b)} \cap S^c|}{|S^c|}
\]
and
\[
\mathrm{SSP}_b(q) =
1\{S \subseteq \widehat S_q^{(b)}\}.
\]
The point plotted in the FPR--SSP plane is
\[
\left(
\frac{1}{B}\sum_{b=1}^B \mathrm{FPR}_b(q),
\frac{1}{B}\sum_{b=1}^B \mathrm{SSP}_b(q)
\right),
\]
where \(B\) is the number of Monte Carlo replications. The labels on selected
points show the corresponding nominal \(q\) level.

# Simulation Setup

The simulation uses fixed sample size
\[
n = `r n`
\]
and number of candidate predictors
\[
p = `r p`.
\]
The true active set is
\[
S = \{1,2,3,4,5\}.
\]
The number of Monte Carlo replications is
\[
B = `r n_rep`.
\]

The q grid is
\[
`r paste(q_levels, collapse = ", ")`.
\]

Predictors are generated independently as
\[
X_{ij} \sim N(0,1).
\]

## Parametric linear model

The parametric model is
\[
Y =
0.22X_1 - 0.20X_2 + 0.18X_3 - 0.16X_4 + 0.14X_5
+ \varepsilon,
\]
with
\[
\varepsilon \sim N(0, 2.3^2).
\]
This is a deliberately weak linear signal, so the curves are not saturated at
\(\mathrm{SSP}=1\).

## Nonparametric nonlinear model

The nonparametric model is
\[
Y = 0.35\{1.6\sin(X_1) + 1.2\tanh(X_2)
+ 0.9X_3^3/(1+X_3^2) - 0.8\exp(0.35X_4) + 0.7X_5\}
+ \sigma(X)\varepsilon,
\]
where
\[
\sigma(X) = 0.7 + 0.5|X_1| + 0.25X_2^2,
\qquad
\varepsilon \sim N(0,0.8^2).
\]
This model is nonlinear and heteroskedastic but retains marginal signal in the
active variables.

# Simulation Results

## FPR versus SSP

```{r fpr-ssp}
ggplot(
  summary_results,
  aes(x = fpr, y = ssp, colour = method, shape = method)
) +
  geom_path(aes(group = method), linewidth = 0.7) +
  geom_point(size = 2.2) +
  geom_text(
    aes(label = q_label),
    size = 2.5,
    nudge_y = 0.025,
    show.legend = FALSE
  ) +
  facet_wrap(~ model) +
  coord_cartesian(xlim = c(0, max(summary_results$fpr) * 1.08),
                  ylim = c(0, 1.04)) +
  labs(
    x = "False positive rate",
    y = "Sure screening property",
    colour = "Method",
    shape = "Method",
    title = "FPR versus SSP across nominal q levels",
    subtitle = paste0("n = ", n, ", p = ", p,
                      ", active variables = ", s,
                      ", replications = ", n_rep)
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

In the parametric linear model, `mfscreen` and Pearson SIS are very close, as
expected: both are driven by marginal linear association. Spearman SIS is less
efficient here because ranking discards some linear Gaussian information.

In the nonlinear heteroskedastic model, all three methods retain the active
variables once \(q\) is moderately large. At smaller \(q\) values, the
model-free statistic remains competitive with Pearson and Spearman while
keeping empirical FPR close to the nominal level. The main advantage of
`mfscreen` is that it uses the reciprocal robust statistic rather than the
ordinary homoskedastic correlation p-value.

## FPR as a function of q

```{r fpr-q}
ggplot(
  summary_results,
  aes(x = q, y = fpr, colour = method, shape = method)
) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ model) +
  scale_x_log10(breaks = q_levels) +
  labs(
    x = "Nominal q level",
    y = "False positive rate",
    colour = "Method",
    shape = "Method",
    title = "False positive rate by q"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

Across both data-generating mechanisms, the empirical FPR follows the nominal
level closely for all three methods. This means that differences in the
FPR--SSP plot are primarily differences in sure screening probability rather
than failures to control false positives.

This behavior is expected in the baseline experiment. The inactive predictors
are independent Gaussian variables, so the two competitors are used in a
setting close to the assumptions behind their usual two-sided correlation
p-values. In this favorable case, plotting FPR as a function of \(q\) mostly
checks that the simulation code and thresholding rules are calibrated.

## FPR-control stress test

The preceding baseline does not strongly challenge Pearson or Spearman
p-values. To probe FPR control more directly, we also run a global-null stress
test with common volatility. In this experiment there are no active variables:
the response and all predictors have zero mean association. However, each row
shares a random volatility factor
\[
A_i = \exp(0.5 Z_i), \qquad Z_i \sim N(0,1),
\]
and
\[
X_{ij} = A_i E_{ij}, \qquad Y_i = A_i \varepsilon_i,
\]
with independent standard normal \(E_{ij}\) and \(\varepsilon_i\).

The ordinary Pearson p-value treats observations as homoskedastic and therefore
uses the wrong null calibration. Spearman ranks reduce sensitivity to the
largest values, but the shared volatility still induces dependence in the
joint ranks. The robust reciprocal statistic in `mfscreen` is designed to
studentize the marginal slope with a heteroskedasticity-aware denominator, so
we expect it to remain much closer to the nominal false positive rate.

```{r stress-fpr}
ggplot(
  stress_results,
  aes(x = q, y = fpr, colour = method, shape = method)
) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey45") +
  geom_line() +
  geom_point(size = 2) +
  scale_x_log10(breaks = q_levels) +
  scale_y_log10() +
  labs(
    x = "Nominal q level",
    y = "Empirical false positive rate",
    colour = "Method",
    shape = "Method",
    title = "FPR control under common volatility",
    subtitle = paste0("Global null, n = ", n, ", p = ", p,
                      ", replications = ", stress_n_rep)
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

The dashed line is the ideal calibration \(\mathrm{FPR}=q\). In this stress
test, `mfscreen` stays close to the dashed line across the q grid. Pearson SIS
is strongly anti-conservative: for example, at \(q=0.05\), its empirical FPR is
about 0.225. Spearman SIS is better than Pearson but still inflated: at
\(q=0.05\), its empirical FPR is about 0.073. The corresponding `mfscreen`
FPR is about 0.054.

```{r stress-inflation}
ggplot(
  stress_results,
  aes(x = q, y = inflation, colour = method, shape = method)
) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey45") +
  geom_line() +
  geom_point(size = 2) +
  scale_x_log10(breaks = q_levels) +
  labs(
    x = "Nominal q level",
    y = "FPR / q",
    colour = "Method",
    shape = "Method",
    title = "False-positive inflation under common volatility"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

The inflation plot makes the calibration gap clearer. Values near one indicate
good FPR control. `mfscreen` is close to one, while the two competitors are
above one throughout the grid, especially Pearson SIS at small \(q\).

## SSP as a function of q

```{r ssp-q}
ggplot(
  summary_results,
  aes(x = q, y = ssp, colour = method, shape = method)
) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ model) +
  scale_x_log10(breaks = q_levels) +
  coord_cartesian(ylim = c(0, 1.04)) +
  labs(
    x = "Nominal q level",
    y = "Sure screening property",
    colour = "Method",
    shape = "Method",
    title = "Sure screening property by q"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

The SSP curves show the tradeoff directly: increasing \(q\) admits more false
positives but improves the probability of retaining all active variables. In
the linear model, `mfscreen` and Pearson SIS overlap almost exactly. In the
nonlinear model, the methods are also close at this sample size, with
`mfscreen` competitive across the full q grid and reaching SSP equal to one
once \(q \geq 0.10\).

# Numerical Summary

```{r table}
summary_table <- summary_results[order(summary_results$model,
                                       summary_results$q,
                                       summary_results$method),
                                 c("model", "q", "method", "fpr", "ssp")]
knitr::kable(
  summary_table,
  digits = 4,
  row.names = FALSE,
  caption = "Mean FPR and SSP over Monte Carlo replications."
)
```
