---
title: "Simulation framework for Table 1"
author: "J. Dedecker, M.-L. Taupin, V. Runge and A.-S. Tocquet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document:
    toc: true
    toc_depth: 2
---

# Objective

This document implements the continuous, heteroscedastic Table 1
data-generating process with a faster simulator for the correlated Gaussian
blocks.

The improvement replaces dense covariance-matrix construction and Cholesky
factorization by the exact stationary Gaussian AR(1) recursion

\[
X_j = \rho X_{j-1} + \sqrt{1-\rho^2}\,\varepsilon_j,
\qquad
\varepsilon_j \overset{i.i.d.}{\sim} \mathcal N(0,\sigma^2),
\]

which gives

\[
\operatorname{Cov}(X_j,X_k)=\sigma^2\rho^{|j-k|}.
\]

Thus the simulated covariance is the same Toeplitz covariance as in Table 1,
but the large \(1982\)-variable noise block is generated much faster.

# Packages

```{r packages}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, cache.path = "table1_cache/")
library(parallel)
```

# Global parameters

```{r parameters}
# Table 1 dimensions
p <- 2000L
N <- 500L

#set.seed(20260701)
base_seed <- 20260701L
replication_seeds <- base_seed + seq_len(N)

n_values <- c(500L, 1000L, 2000L, 2500L)

# Variable groups
n_active <- 10L
n_proxy_and_noise <- 8L
n_toeplitz_noise <- p - n_active - n_proxy_and_noise  # 1982

# Toeplitz correlations
rho_active <- 0.57
rho_toeplitz_noise <- 0.70

# Regression coefficients, including the intercept
theta_star <- c(
  -1,
   3.619350,
  -3.274923,
   2.963273,
  -2.681280,
   2, 4, 6, 3, 2, 4
)

stopifnot(length(theta_star) == n_active + 1L)
stopifnot(n_toeplitz_noise == 1982L)
```

# Fast Toeplitz-Gaussian generator

```{r fast-toeplitz}
simulate_toeplitz_gaussian_fast <- function(n, d, rho, sigma = 1) 
{
  stopifnot(d >= 1L, abs(rho) < 1, sigma > 0)

  X <- matrix(0, nrow = n, ncol = d)

  # A stationary AR(1) process starts with marginal variance sigma^2.
  X[, 1L] <- rnorm(n, mean = 0, sd = sigma)

  if (d > 1L) {
    innovation_sd <- sigma * sqrt(1 - rho^2)

    # Draw all innovations at once to reduce random-number-generation overhead.
    innovations <- matrix(
      rnorm(n * (d - 1L), mean = 0, sd = innovation_sd),
      nrow = n,
      ncol = d - 1L
    )

    # Each row is a stationary Gaussian AR(1) sequence.
    for (j in 2:d) {
      X[, j] <- rho * X[, j - 1L] + innovations[, j - 1L]
    }
  }

  X
}
```

# One Table 1 data set

This function returns only \(X\) and \(Y\) by default. It does not create
column names or duplicate metadata at every replication, reducing memory
allocation in Monte Carlo experiments.

```{r one-dataset}
simulate_table1_dataset_fast <- function(
    n,
    p = 2000,
    rho_active = 0.57,
    rho_toeplitz_noise = 0.70,
    theta_star = c(-1, 3.619350, -3.274923, 2.963273, -2.681280, 2, 4, 6, 3,2, 4),
    add_names = FALSE
) {
  # ---------------------------------------------------------------
  # Group A: X1,...,X10
  # ---------------------------------------------------------------

  X1_to_X5 <- simulate_toeplitz_gaussian_fast(
    n = n,
    d = 5L,
    rho = rho_active,
    sigma = 1
  )

  X6 <- rbinom(n, size = 1L, prob = 0.35)
  X7 <- rchisq(n, df = 2)
  X8 <- X1_to_X5[, 1L] * rpois(n, lambda = 2)
  X9 <- X1_to_X5[, 2L] * rnorm(n, mean = 1, sd = 1)
  X10 <- rt(n, df = 14)

  X_active <- cbind(X1_to_X5, X6, X7, X8, X9, X10)

  # ---------------------------------------------------------------
  # Response Y
  # ---------------------------------------------------------------

  linear_predictor <- drop(theta_star[1L] + X_active %*% theta_star[-1L])
  epsilon <- X_active[, 5L] * (rchisq(n, df = 3) - 3)
  Y <- abs(linear_predictor)^0.8 + epsilon / 3

  # ---------------------------------------------------------------
  # Groups B and C: X11,...,X18
  # ---------------------------------------------------------------

  Z3 <- rnorm(n, mean = 0, sd = 0.40)
  Z4 <- rnorm(n, mean = 0, sd = 0.02)
  Z5 <- rnorm(n, mean = 0, sd = 0.35)
  Z6 <- rnorm(n, mean = 0, sd = 0.55)

  X_proxy_and_noise <- cbind(
    X_active[, 1L] + Z3,  # X11
    X_active[, 3L] + Z4,  # X12
    X_active[, 4L] + Z5,  # X13
    X_active[, 5L] + Z6,  # X14
    Z3, Z4, Z5, Z6        # X15,...,X18
  )

  # ---------------------------------------------------------------
  # Group D: X19,...,X2000
  # ---------------------------------------------------------------

  X19_to_X2000 <- simulate_toeplitz_gaussian_fast(
    n = n,
    d = p - 18L,
    rho = rho_toeplitz_noise,
    sigma = 1
  )

  X <- cbind(X_active, X_proxy_and_noise, X19_to_X2000)

  stopifnot(ncol(X) == p)
  stopifnot(length(Y) == n)

  if (add_names) {
    colnames(X) <- paste0("X", seq_len(p))
  }

  list(X = X, Y = Y)
}
```

# Shared metadata

Keep this object once, outside Monte Carlo replications.

```{r metadata}
table1_metadata <- list(
  p = p,
  N = N,
  n_values = n_values,
  theta_star = theta_star,
  rho_active = rho_active,
  rho_toeplitz_noise = rho_toeplitz_noise,
  active_indices = 1:10,
  proxy_indices = 11:14,
  null_noise_indices = 15:18,
  toeplitz_noise_indices = 19:p,
  table1_positive_indices = 1:14,
  table1_null_indices = 15:p
)
```

# Example data set

```{r example}
example_dataset <- simulate_table1_dataset_fast(
  n = 500L,
    add_names = FALSE
)

dim(example_dataset$X)
head(example_dataset$X[, 1:12])
head(example_dataset$Y)

# Empirical correlation of X1,...,X5, to inspect the rho = 0.57 block:
round(cor(example_dataset$X[, 1:5]), 3)
# Empirical correlation of X150,...,X155, to inspect the rho = 0.7 block:
round(cor(example_dataset$X[, 150:155]), 3)
```



# Screening function from the `mfscreen` package

The screening routine is provided by the GitHub package
[`vrunge/mfscreen`](https://github.com/vrunge/mfscreen). The following chunk
installs `remotes` only when needed, installs `mfscreen` only when it is not
already available, and then loads the package.

```{r install-mfscreen, message=TRUE, warning=TRUE}
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}

if (!requireNamespace("mfscreen", quietly = TRUE)) {
  remotes::install_github("vrunge/mfscreen")
}

library(mfscreen)

# Confirm that the required function is available.
stopifnot(exists("screening_test_matrix", where = asNamespace("mfscreen")))
```

# One replication: data generation and screening

This function generates one Table 1 data set, immediately applies the Rcpp
screening function, and returns only the quantities needed for Table 1.
The \(n \times p\) design matrix is not retained after the function returns.

```{r one-replication}
run_one_table1_replication <- function(n, q = 0.10) {
  dat <- simulate_table1_dataset_fast(
    n = n,
    p = p,
    rho_active = rho_active,
    rho_toeplitz_noise = rho_toeplitz_noise,
    theta_star = theta_star,
    add_names = FALSE
  )

  test_result <- mfscreen::screening_test_matrix(
    X = dat$X,
    y = as.numeric(dat$Y),
    q = q
  )

  selected <- as.logical(test_result$selected)

  # Table 1: X1,...,X14 are treated as positive; X15,...,X2000 are null.
  selected_first18 <- as.integer(selected[1:18])

  output <- list(
    selected_first18 = selected_first18,
    selected_positive = sum(selected[1:14]),
    selected_null = sum(selected[15:p]),
    selected_total = sum(selected),
    threshold = test_result$threshold
  )

  output
}
```

# Example: one replication

```{r one-replication-example}
one_result <- run_one_table1_replication(n = 500L, q = 0.10)

one_replication_summary <- data.frame(
  n = 500L,
  p = p,
  q = 0.10,
  threshold = one_result$threshold,
  selected_among_X1_to_X14 = one_result$selected_positive,
  selected_among_X15_to_X2000 = one_result$selected_null,
  selected_model_size = one_result$selected_total,
  TPR_contribution = one_result$selected_positive / 14,
  FPR_contribution = one_result$selected_null / (p - 14)
)

knitr::kable(
  one_replication_summary,
  digits = 4,
  caption = "One replication: quantities that are averaged for Table 1"
)

knitr::kable(
  data.frame(
    variable = paste0("X", 1:18),
    selected = one_result$selected_first18
  ),
  caption = "One replication: selection indicators for X1 to X18"
)
```

# Monte Carlo simulation for one Table 1 column

The function below runs \(N\) replications for a fixed sample size and returns
only compact outputs. It does not save any \(X\) or \(Y\) matrices.

```{r monte-carlo-function}

run_table1_column <- function(n, N = N, q = 0.10, 
                              mc.cores = 1L,
                              base_seed = 20260701L)
{

  replication_seeds <- base_seed + seq_len(N)

  one_draw <- function(i) {
    set.seed(replication_seeds[i])
    run_one_table1_replication(n = n, q = q)
  }

  if (.Platform$OS.type == "windows" || mc.cores <= 1L) {
    repetitions <- lapply(seq_len(N), one_draw)
  } else {
    repetitions <- parallel::mclapply(
      X = seq_len(N),
      FUN = one_draw,
      mc.cores = mc.cores,
      mc.set.seed = FALSE,
      mc.preschedule = FALSE
    )
  }

  selection_first18 <- do.call(
    rbind,
    lapply(repetitions, `[[`, "selected_first18")
  )

  selected_positive <- vapply(
    repetitions,
    `[[`,
    numeric(1),
    "selected_positive"
  )

  selected_null <- vapply(
    repetitions,
    `[[`,
    numeric(1),
    "selected_null"
  )

  selected_total <- vapply(
    repetitions,
    `[[`,
    numeric(1),
    "selected_total"
  )

  list(
    n = n,
    N = N,
    q = q,
    threshold = repetitions[[1]]$threshold,
    selection_frequency_first18 = colMeans(selection_first18),
    TPR = mean(selected_positive / 14),
    FPR = mean(selected_null / (p - 14)),
    mean_selected_model_size = mean(selected_total)
  )
}
```

# Generate Table 1

This chunk is evaluated when the document is knitted. It runs the simulation
for \(n=500,1000,2000,2500\), each with \(N=500\) Monte Carlo replications,
and then creates the generated Table 1 in the PDF.

For a quick development check, temporarily replace `N <- 500L` in the
parameter section by a smaller value such as `N <- 5L`.

```{r table1-run, eval=TRUE, cache=TRUE}
# On machines with limited RAM, start with 4 to 8 cores.
n_cores <- min(8L, parallel::detectCores(logical = FALSE))

table1_runs <- lapply(
  n_values,
  run_table1_column,
  N = N,
  q = 0.10,
  mc.cores = n_cores
)

names(table1_runs) <- paste0("n=", n_values)
```

# Generated Table 1

The generated Table 1 reports one row for each sample size and the three
Monte Carlo quantities reported in the paper: TPR, FPR, and the mean selected
model size.

```{r table1-summary, eval=TRUE, cache=TRUE}
table1_summary <- data.frame(
  n = n_values,
  TPR = vapply(table1_runs, `[[`, numeric(1), "TPR"),
  FPR = vapply(table1_runs, `[[`, numeric(1), "FPR"),
  Mean_selected_model_size = vapply(
    table1_runs,
    `[[`,
    numeric(1),
    "mean_selected_model_size"
  )
)

knitr::kable(
  table1_summary,
  format = "latex",
  booktabs = TRUE,
  digits = 3,
  align = c("c", "c", "c", "c"),
  caption = paste0(
    "Generated Table 1: Monte Carlo screening results ",
    "(N = ", N, ", p = ", p, ", q = 0.10)"
  )
)
```

# Selection frequencies for X1 to X18

This table contains the Monte Carlo selection frequencies for the first
18 covariates. The Table 1 TPR is the mean of the first 14 rows; the FPR is
computed over all \(X_{15},\ldots,X_{2000}\), not only \(X_{15},\ldots,X_{18}\).

```{r first18-summary, eval=TRUE, cache=TRUE}
selection_first18_table <- data.frame(
  variable = paste0("X", 1:18),
  do.call(
    cbind,
    lapply(
      table1_runs,
      function(x) x$selection_frequency_first18
    )
  ),
  check.names = FALSE
)

colnames(selection_first18_table)[-1] <- names(table1_runs)

selection_first18_table[, -1] <- lapply(
  selection_first18_table[, -1, drop = FALSE],
  round,
  digits = 3
)

knitr::kable(
  selection_first18_table,
  format = "latex",
  booktabs = TRUE,
  caption = "Monte Carlo selection frequencies for X1 to X18"
)
```

# Save compact results

```{r save-results, eval=TRUE}
saveRDS(
  object = list(
    metadata = table1_metadata,
    table1_summary = table1_summary,
    table1_runs = table1_runs
  ),
  file = "table1_rcpp_screening_results.rds"
)

write.csv(
  table1_summary,
  file = "table1_rcpp_screening_summary.csv",
  row.names = FALSE
)
```
