---
title: "Simulation framework for Table 1 — fast data-generating process"
author: "J. Dedecker, M.-L. Taupin et A.-S. Tocquet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document:
    toc: true
    toc_depth: 2
---

# Objective

This document implements the continuous, heteroscedastic Table 1
data-generating process with a faster simulator for the correlated Gaussian
blocks.

The improvement replaces dense covariance-matrix construction and Cholesky
factorization by the exact stationary Gaussian AR(1) recursion

\[
X_j = \rho X_{j-1} + \sqrt{1-\rho^2}\,\varepsilon_j,
\qquad
\varepsilon_j \overset{i.i.d.}{\sim} \mathcal N(0,\sigma^2),
\]

which gives

\[
\operatorname{Cov}(X_j,X_k)=\sigma^2\rho^{|j-k|}.
\]

Thus the simulated covariance is the same Toeplitz covariance as in Table 1,
but the large \(1982\)-variable noise block is generated much faster.

# Packages

```{r packages}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)

library(parallel)
```

# Global parameters

```{r parameters}
set.seed(20260701)

# Table 1 dimensions
p <- 2000L
N <- 500L
n_values <- c(500L, 1000L, 2000L, 2500L)

# Variable groups
n_active <- 10L
n_proxy_and_noise <- 8L
n_toeplitz_noise <- p - n_active - n_proxy_and_noise  # 1982

# Toeplitz correlations
rho_active <- 0.57
rho_toeplitz_noise <- 0.70

# Regression coefficients, including the intercept
theta_star <- c(
  -1,
   3.619350,
  -3.274923,
   2.963273,
  -2.681280,
   2, 4, 6, 3, 2, 4
)

stopifnot(length(theta_star) == n_active + 1L)
stopifnot(n_toeplitz_noise == 1982L)
```

# Fast Toeplitz-Gaussian generator

```{r fast-toeplitz}
simulate_toeplitz_gaussian_fast <- function(n, d, rho, sigma = 1) {
  stopifnot(d >= 1L, abs(rho) < 1, sigma > 0)

  X <- matrix(0, nrow = n, ncol = d)

  # A stationary AR(1) process starts with marginal variance sigma^2.
  X[, 1L] <- rnorm(n, mean = 0, sd = sigma)

  if (d > 1L) {
    innovation_sd <- sigma * sqrt(1 - rho^2)

    # Draw all innovations at once to reduce random-number-generation overhead.
    innovations <- matrix(
      rnorm(n * (d - 1L), mean = 0, sd = innovation_sd),
      nrow = n,
      ncol = d - 1L
    )

    # Each row is a stationary Gaussian AR(1) sequence.
    for (j in 2:d) {
      X[, j] <- rho * X[, j - 1L] + innovations[, j - 1L]
    }
  }

  X
}
```

# One Table 1 data set

This function returns only \(X\) and \(Y\) by default. It does not create
column names or duplicate metadata at every replication, reducing memory
allocation in Monte Carlo experiments.

```{r one-dataset}
simulate_table1_dataset_fast <- function(
    n,
    p = p,
    rho_active = rho_active,
    rho_toeplitz_noise = rho_toeplitz_noise,
    theta_star = theta_star,
    add_names = FALSE
) {
  # ---------------------------------------------------------------
  # Group A: X1,...,X10
  # ---------------------------------------------------------------

  X1_to_X5 <- simulate_toeplitz_gaussian_fast(
    n = n,
    d = 5L,
    rho = rho_active,
    sigma = 1
  )

  X6 <- rbinom(n, size = 1L, prob = 0.35)
  X7 <- rchisq(n, df = 2)
  X8 <- X1_to_X5[, 1L] * rpois(n, lambda = 2)
  X9 <- X1_to_X5[, 2L] * rnorm(n, mean = 1, sd = 1)
  X10 <- rt(n, df = 14)

  X_active <- cbind(X1_to_X5, X6, X7, X8, X9, X10)

  # ---------------------------------------------------------------
  # Response Y
  # ---------------------------------------------------------------

  linear_predictor <- drop(theta_star[1L] + X_active %*% theta_star[-1L])
  epsilon <- X_active[, 5L] * (rchisq(n, df = 3) - 3)
  Y <- abs(linear_predictor)^0.8 + epsilon / 3

  # ---------------------------------------------------------------
  # Groups B and C: X11,...,X18
  # ---------------------------------------------------------------

  Z3 <- rnorm(n, mean = 0, sd = 0.40)
  Z4 <- rnorm(n, mean = 0, sd = 0.02)
  Z5 <- rnorm(n, mean = 0, sd = 0.35)
  Z6 <- rnorm(n, mean = 0, sd = 0.55)

  X_proxy_and_noise <- cbind(
    X_active[, 1L] + Z3,  # X11
    X_active[, 3L] + Z4,  # X12
    X_active[, 4L] + Z5,  # X13
    X_active[, 5L] + Z6,  # X14
    Z3, Z4, Z5, Z6        # X15,...,X18
  )

  # ---------------------------------------------------------------
  # Group D: X19,...,X2000
  # ---------------------------------------------------------------

  X19_to_X2000 <- simulate_toeplitz_gaussian_fast(
    n = n,
    d = p - 18L,
    rho = rho_toeplitz_noise,
    sigma = 1
  )

  X <- cbind(X_active, X_proxy_and_noise, X19_to_X2000)

  stopifnot(ncol(X) == p)
  stopifnot(length(Y) == n)

  if (add_names) {
    colnames(X) <- paste0("X", seq_len(p))
  }

  list(X = X, Y = Y)
}
```

# Shared metadata

Keep this object once, outside Monte Carlo replications.

```{r metadata}
table1_metadata <- list(
  p = p,
  N = N,
  n_values = n_values,
  theta_star = theta_star,
  rho_active = rho_active,
  rho_toeplitz_noise = rho_toeplitz_noise,
  active_indices = 1:10,
  proxy_indices = 11:14,
  null_noise_indices = 15:18,
  toeplitz_noise_indices = 19:p,
  table1_positive_indices = 1:14,
  table1_null_indices = 15:p
)
```

# Example data set

```{r example}
example_dataset <- simulate_table1_dataset_fast(
  n = 500L,
    p = p,
    rho_active = rho_active,
    rho_toeplitz_noise = rho_toeplitz_noise,
    theta_star = theta_star,
    add_names = FALSE
)

dim(example_dataset$X)
head(example_dataset$X[, 1:12])
head(example_dataset$Y)

# Empirical correlation of X1,...,X5, to inspect the rho = 0.57 block:
round(cor(example_dataset$X[, 1:5]), 3)
# Empirical correlation of X150,...,X155, to inspect the rho = 0.7 block:
round(cor(example_dataset$X[, 150:155]), 3)
```

# Monte Carlo framework

Use the following only when you need all simulated data sets. It is memory
intensive for large \(n\) and large \(N\).

```{r monte-carlo, eval=FALSE}
simulate_table1_monte_carlo_fast <- function(
    n,
    N = N,
    mc.cores = 1L
) {
  one_draw <- function(i) {
    simulate_table1_dataset_fast(n = n, add_names = FALSE)
  }

  if (.Platform$OS.type == "windows" || mc.cores <= 1L) {
    lapply(seq_len(N), one_draw)
  } else {
    parallel::mclapply(
      X = seq_len(N),
      FUN = one_draw,
      mc.cores = mc.cores,
      mc.set.seed = TRUE,
      mc.preschedule = FALSE
    )
  }
}
```

# Recommended workflow for a screening study

When screening will immediately follow data generation, do not store all
\(X\) and \(Y\) matrices. Compute the screening output inside each
replication and return only the compact statistic or selected indices.

```{r compact-workflow, eval=FALSE}
one_draw_returning_summary <- function(i, n) {
  dat <- simulate_table1_dataset_fast(n = n, add_names = FALSE)

  # Replace this placeholder with the desired screening procedure.
  # result <- screening_test_matrix(dat$X, dat$Y, q = 0.10)

  # Return only compact output, for example:
  # list(selected_indices = result$selected_indices)
}

n_cores <- min(8L, parallel::detectCores(logical = FALSE))

# summaries <- parallel::mclapply(
#   seq_len(N),
#   one_draw_returning_summary,
#   n = 500L,
#   mc.cores = n_cores,
#   mc.set.seed = TRUE,
#   mc.preschedule = FALSE
# )
```
