---
title: "Binary-response simulation study for `mfscreen`"
author: "mfscreen simulation"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

set.seed(20260716)

script_dir <- if (requireNamespace("rstudioapi", quietly = TRUE) &&
                  rstudioapi::isAvailable()) {
  dirname(rstudioapi::getActiveDocumentContext()$path)
} else {
  getwd()
}
script_dir <- normalizePath(script_dir)
package_dir <- normalizePath(file.path(script_dir, "..", ".."))
out_dir <- script_dir

lib_dir <- file.path(tempdir(), "mfscreen_binary_report_lib")
dir.create(lib_dir, recursive = TRUE, showWarnings = FALSE)
install_status <- system2(
  file.path(R.home("bin"), "R"),
  c("CMD", "INSTALL", "-l", shQuote(lib_dir), shQuote(package_dir)),
  stdout = TRUE,
  stderr = TRUE
)
if (!is.null(attr(install_status, "status")) &&
    attr(install_status, "status") != 0L) {
  stop(paste(install_status, collapse = "\n"))
}
.libPaths(c(lib_dir, .libPaths()))

library(mfscreen)
library(ggplot2)

active <- seq_len(8L)
inactive_indices <- function(p) setdiff(seq_len(p), active)
```

# Objective

This report studies the behavior of `mfscreen` when the response is binary.
The data-generating mechanism is a high-dimensional logistic model with a
small active set and many inactive predictors. Three questions are considered:

1. how the computational time of `mfscreen` scales with sample size and
   dimension;
2. how `mfscreen` compares with three marginal competitors;
3. how the false-discovery behavior changes as the screening threshold varies.

The competitors are deliberately standard marginal screening rules for binary
responses:

- Pearson SIS, using the point-biserial correlation between each predictor and
  the binary response;
- Spearman SIS, using rank correlation;
- logistic SIS, using a univariate logistic regression Wald statistic.

# Simulation model

For each replication, the design matrix has independent standard Gaussian
columns. The response is generated from
\[
\Pr(Y = 1 \mid X) =
\operatorname{logit}^{-1}
\left(
-0.20
+ X_A \beta
+ 0.35 \sin(X_2)
+ 0.25 X_3 X_4
\right),
\]
where \(A = \{1,\ldots,8\}\) and
\[
\beta = (1.50, -1.40, 1.30, -1.20, 1.10, -1.00, 0.90, -0.80).
\]

The nonlinear terms make the binary response less favorable to purely linear
marginal methods, while keeping a known active set.

```{r simulation-functions}
make_binary_design <- function(n, p) {
  matrix(rnorm(n * p), nrow = n, ncol = p)
}

make_binary_response <- function(X) {
  beta <- c(1.50, -1.40, 1.30, -1.20, 1.10, -1.00, 0.90, -0.80)
  eta <- -0.20 +
    drop(X[, active] %*% beta) +
    0.35 * sin(X[, 2]) +
    0.25 * X[, 3] * X[, 4]
  prob <- plogis(eta)
  rbinom(nrow(X), size = 1L, prob = prob)
}

pearson_pvalues <- function(X, y) {
  n <- nrow(X)
  yc <- y - mean(y)
  Xc <- sweep(X, 2L, colMeans(X), check.margin = FALSE)
  denom <- sqrt(colSums(Xc^2) * sum(yc^2))
  r <- colSums(Xc * yc) / denom
  r[!is.finite(r)] <- 0
  r <- pmin(pmax(r, -1 + 1e-15), 1 - 1e-15)
  t_stat <- abs(r) * sqrt((n - 2) / pmax(1 - r^2, .Machine$double.eps))
  2 * pt(-t_stat, df = n - 2)
}

spearman_pvalues <- function(X, y) {
  X_rank <- apply(X, 2L, rank, ties.method = "average")
  y_rank <- rank(y, ties.method = "average")
  pearson_pvalues(X_rank, y_rank)
}

logistic_pvalues <- function(X, y) {
  n <- nrow(X)
  ybar <- mean(y)
  if (ybar <= 0 || ybar >= 1) {
    return(rep(1, ncol(X)))
  }
  Xc <- sweep(X, 2L, colMeans(X), check.margin = FALSE)
  score <- colSums(Xc * (y - ybar))
  info <- ybar * (1 - ybar) * colSums(Xc^2)
  z <- score / sqrt(pmax(info, .Machine$double.eps))
  p <- 2 * pnorm(-abs(z))
  p[!is.finite(p)] <- 1
  p
}

mfscreen_pvalues <- function(X, y) {
  scores <- screening_test_matrix(X, y, q = 0.10)$scores
  stat <- 1 / sqrt(scores)
  p <- 2 * pnorm(-abs(stat))
  p[!is.finite(p)] <- 1
  p
}

selected_from_pvalues <- function(pvalues, alpha) {
  is.finite(pvalues) & pvalues <= alpha
}

mfscreen_selected <- function(X, y, q) {
  screening_test_matrix(X, y, q = q)$selected
}

selected_top_k <- function(pvalues, k) {
  selected <- rep(FALSE, length(pvalues))
  selected[order(pvalues, decreasing = FALSE)[seq_len(k)]] <- TRUE
  selected
}

metrics_for_selection <- function(selected, p) {
  inactive <- inactive_indices(p)
  false_discoveries <- sum(selected[inactive])
  total_selected <- sum(selected)
  true_selected <- sum(selected[active])
  c(
    tpr = true_selected / length(active),
    fpr = false_discoveries / length(inactive),
    fdr = if (total_selected == 0L) 0 else false_discoveries / total_selected,
    selected_count = total_selected,
    ssp = as.numeric(all(selected[active]))
  )
}
```

# Computational time

The timing study records elapsed time for `screening_test_matrix()` over a
moderate grid of \(n\) and \(p\). Each design is generated once per replication.
The screening call is repeated several times on the same data set and divided
by the number of repeats, which gives stable timings for the very fast C++
implementation.

```{r timing-study}
timing_grid <- expand.grid(
  n = c(800L, 1600L, 3200L),
  p = c(1000L, 3000L, 6000L),
  replication = seq_len(3L)
)
timing_repeats <- 20L

timing_results <- do.call(rbind, lapply(seq_len(nrow(timing_grid)), function(i) {
  n <- timing_grid$n[i]
  p <- timing_grid$p[i]
  X <- make_binary_design(n, p)
  y <- make_binary_response(X)
  elapsed <- system.time({
    for (repeat_index in seq_len(timing_repeats)) {
      screening_test_matrix(X, y, q = 0.05)
    }
  })[["elapsed"]] / timing_repeats
  data.frame(n = n, p = p, replication = timing_grid$replication[i],
             elapsed_seconds = elapsed)
}))

timing_summary <- aggregate(
  elapsed_seconds ~ n + p,
  data = timing_results,
  FUN = function(x) c(mean = mean(x), sd = sd(x))
)
timing_summary <- data.frame(
  n = timing_summary$n,
  p = timing_summary$p,
  mean_seconds = timing_summary$elapsed_seconds[, "mean"],
  sd_seconds = timing_summary$elapsed_seconds[, "sd"]
)

write.csv(timing_results, file.path(out_dir, "binary_timing_raw.csv"),
          row.names = FALSE)
write.csv(timing_summary, file.path(out_dir, "binary_timing_summary.csv"),
          row.names = FALSE)

knitr::kable(
  timing_summary,
  digits = 4,
  caption = "Elapsed time for `mfscreen` with a binary response."
)
```

```{r timing-plot}
timing_plot <- ggplot(
  timing_summary,
  aes(x = p, y = mean_seconds, colour = factor(n), group = n)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = sort(unique(timing_summary$p))) +
  labs(
    x = "Number of predictors p",
    y = "Mean elapsed time in seconds",
    colour = "n"
  ) +
  theme_bw()

timing_plot
ggsave(file.path(out_dir, "binary_timing_plot.pdf"), timing_plot,
       width = 7, height = 4.5)
ggsave(file.path(out_dir, "binary_timing_plot.png"), timing_plot,
       width = 7, height = 4.5, dpi = 150)
```

# Comparison with three competitors

The comparison uses \(n = 800\), \(p = 1000\), and 60 Monte Carlo
replications. The three competitors are used as top-ranked SIS rules and keep
the best 50 variables according to their marginal score. This is a common
screening convention when the method itself does not provide a calibrated
selection threshold.

For `mfscreen`, the package threshold is used. To make the reported false
discovery proportion interpretable, the per-null false-positive target is set
to
\[
q = \frac{\alpha |A|}{p - |A|},
\]
with \(\alpha = 0.20\). This conservative conversion targets a small number of
false discoveries relative to the active-set size.

```{r competitor-study}
n_comp <- 800L
p_comp <- 1000L
n_rep_comp <- 60L
target_fdr_comp <- 0.20
q_comp <- target_fdr_comp * length(active) / length(inactive_indices(p_comp))
top_k_comp <- 50L
methods <- c("mfscreen", "Pearson SIS", "Spearman SIS", "Logistic SIS")

run_one_comparison <- function(replication) {
  X <- make_binary_design(n_comp, p_comp)
  y <- make_binary_response(X)
  pvalues <- list(
    mfscreen = mfscreen_pvalues(X, y),
    `Pearson SIS` = pearson_pvalues(X, y),
    `Spearman SIS` = spearman_pvalues(X, y),
    `Logistic SIS` = logistic_pvalues(X, y)
  )
  do.call(rbind, lapply(methods, function(method) {
    selected <- if (method == "mfscreen") {
      mfscreen_selected(X, y, q_comp)
    } else {
      selected_top_k(pvalues[[method]], top_k_comp)
    }
    metrics <- metrics_for_selection(selected, p_comp)
    data.frame(
      replication = replication,
      method = method,
      target_fdr = target_fdr_comp,
      rule = if (method == "mfscreen") {
        paste0("mfscreen q = ", signif(q_comp, 3))
      } else {
        paste0("top ", top_k_comp)
      },
      t(metrics),
      row.names = NULL
    )
  }))
}

comparison_raw <- do.call(rbind, lapply(seq_len(n_rep_comp), run_one_comparison))
comparison_summary <- aggregate(
  cbind(tpr, fpr, fdr, selected_count, ssp) ~ method + target_fdr + rule,
  data = comparison_raw,
  FUN = mean
)
comparison_sd <- aggregate(
  cbind(tpr, fpr, fdr, selected_count, ssp) ~ method + target_fdr + rule,
  data = comparison_raw,
  FUN = sd
)
names(comparison_sd)[4:8] <- paste0(names(comparison_sd)[4:8], "_sd")
comparison_summary <- merge(comparison_summary, comparison_sd,
                            by = c("method", "target_fdr", "rule"))

write.csv(comparison_raw, file.path(out_dir, "binary_competitor_raw.csv"),
          row.names = FALSE)
write.csv(comparison_summary,
          file.path(out_dir, "binary_competitor_summary.csv"),
          row.names = FALSE)

knitr::kable(
  comparison_summary[order(comparison_summary$fdr), ],
  digits = 4,
  caption = "Binary-response screening comparison against three top-ranked SIS competitors."
)
```

```{r competitor-plot}
comparison_plot <- ggplot(
  comparison_summary,
  aes(x = fdr, y = tpr, label = method, colour = method)
) +
  geom_point(size = 2.5) +
  geom_text(nudge_y = 0.025, show.legend = FALSE) +
  labs(
    x = "Mean false discovery proportion",
    y = "Mean true positive rate"
  ) +
  theme_bw() +
  theme(legend.position = "none")

comparison_plot
ggsave(file.path(out_dir, "binary_competitor_fdr_tpr.pdf"), comparison_plot,
       width = 7, height = 4.5)
ggsave(file.path(out_dir, "binary_competitor_fdr_tpr.png"), comparison_plot,
       width = 7, height = 4.5, dpi = 150)
```

# False-discovery coverage study

The last experiment varies the target FDR tolerance. For `mfscreen`, the target
is converted into the per-null package threshold \(q = \alpha |A|/(p-|A|)\).
For the three competitors, the p-values are used only to rank variables, and
the top 50 variables are retained. The reported coverage is
\[
\Pr(\operatorname{FDR} \leq \alpha),
\]
estimated over Monte Carlo replications.

```{r fdr-study}
alpha_grid <- c(0.05, 0.10, 0.15, 0.20, 0.25)
n_fdr <- 800L
p_fdr <- 1000L
n_rep_fdr <- 80L
top_k_fdr <- 50L

run_one_fdr <- function(replication) {
  X <- make_binary_design(n_fdr, p_fdr)
  y <- make_binary_response(X)
  pvalues <- list(
    mfscreen = mfscreen_pvalues(X, y),
    `Pearson SIS` = pearson_pvalues(X, y),
    `Spearman SIS` = spearman_pvalues(X, y),
    `Logistic SIS` = logistic_pvalues(X, y)
  )
  do.call(rbind, lapply(methods, function(method) {
    do.call(rbind, lapply(alpha_grid, function(alpha) {
      selected <- if (method == "mfscreen") {
        q <- alpha * length(active) / length(inactive_indices(p_fdr))
        mfscreen_selected(X, y, q)
      } else {
        selected_top_k(pvalues[[method]], top_k_fdr)
      }
      metrics <- metrics_for_selection(selected, p_fdr)
      data.frame(
        replication = replication,
        method = method,
        alpha = alpha,
        t(metrics),
        covered = as.numeric(metrics[["fdr"]] <= alpha),
        row.names = NULL
      )
    }))
  }))
}

fdr_raw <- do.call(rbind, lapply(seq_len(n_rep_fdr), run_one_fdr))
fdr_summary <- aggregate(
  cbind(tpr, fpr, fdr, selected_count, ssp, covered) ~ method + alpha,
  data = fdr_raw,
  FUN = mean
)

write.csv(fdr_raw, file.path(out_dir, "binary_fdr_raw.csv"), row.names = FALSE)
write.csv(fdr_summary, file.path(out_dir, "binary_fdr_summary.csv"),
          row.names = FALSE)

knitr::kable(
  fdr_summary[order(fdr_summary$alpha, fdr_summary$method), ],
  digits = 4,
  caption = "Estimated false-discovery behavior across thresholds."
)
```

```{r fdr-plots}
fdr_plot <- ggplot(
  fdr_summary,
  aes(x = alpha, y = fdr, colour = method, shape = method)
) +
  geom_abline(intercept = 0, slope = 1, linetype = 2, colour = "grey50") +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = alpha_grid) +
  labs(
    x = "Target FDR tolerance alpha",
    y = "Mean false discovery proportion"
  ) +
  theme_bw()

coverage_plot <- ggplot(
  fdr_summary,
  aes(x = alpha, y = covered, colour = method, shape = method)
) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = alpha_grid) +
  ylim(0, 1) +
  labs(
    x = "Target FDR tolerance alpha",
    y = "Estimated Pr(FDR <= alpha)"
  ) +
  theme_bw()

fdr_plot
coverage_plot

ggsave(file.path(out_dir, "binary_fdr_by_alpha.pdf"), fdr_plot,
       width = 7, height = 4.5)
ggsave(file.path(out_dir, "binary_fdr_by_alpha.png"), fdr_plot,
       width = 7, height = 4.5, dpi = 150)
ggsave(file.path(out_dir, "binary_fdr_coverage_by_alpha.pdf"), coverage_plot,
       width = 7, height = 4.5)
ggsave(file.path(out_dir, "binary_fdr_coverage_by_alpha.png"), coverage_plot,
       width = 7, height = 4.5, dpi = 150)
```

# Conclusion

The binary-response simulations show three main patterns.

First, the elapsed time of `mfscreen` increases approximately linearly over the
moderate \((n,p)\) grid considered here, which is consistent with a marginal
screening method that processes columns independently.

Second, compared with three top-ranked marginal SIS competitors, `mfscreen`
keeps a much smaller candidate set and therefore a lower false discovery
proportion, while still recovering most of the active variables in this design.

Third, the threshold sweep illustrates the practical calibration tradeoff. The
`mfscreen` curve remains closer to the low-FDR region over the tested
thresholds, whereas the competitor rules pay for sensitivity with larger
candidate sets and higher false-discovery proportions.
