---
title: "ZIP-response screening with the ZIP simulation design"
author: "mfscreen simulation"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7.2,
  fig.height = 4.8
)

set.seed(20260716)

script_dir <- if (requireNamespace("rstudioapi", quietly = TRUE) &&
                  rstudioapi::isAvailable()) {
  dirname(rstudioapi::getActiveDocumentContext()$path)
} else {
  getwd()
}
script_dir <- normalizePath(script_dir)
package_dir <- normalizePath(file.path(script_dir, "..", ".."))
out_dir <- script_dir

lib_dir <- file.path(tempdir(), "mfscreen_zip_response_report_lib")
dir.create(lib_dir, recursive = TRUE, showWarnings = FALSE)
install_status <- system2(
  file.path(R.home("bin"), "R"),
  c("CMD", "INSTALL", "-l", shQuote(lib_dir), shQuote(package_dir)),
  stdout = TRUE,
  stderr = TRUE
)
if (!is.null(attr(install_status, "status")) &&
    attr(install_status, "status") != 0L) {
  stop(paste(install_status, collapse = "\n"))
}
.libPaths(c(lib_dir, .libPaths()))

library(mfscreen)
library(ggplot2)

theme_set(theme_bw(base_size = 11))
method_colours <- c(
  "mfscreen" = "#1B9E77",
  "Pearson SIS" = "#D95F02",
  "Spearman SIS" = "#7570B3",
  "Poisson-score SIS" = "#E7298A",
  "Poisson-Wald SIS" = "#66A61E"
)
```

# Objective

This report reproduces the simulation framework of
`3_ZIP_abs_Simulations_screening_Version pour Vincent.Rmd`. The response is
the original zero-inflated Poisson response
\[
Y = Y_1Y_2,
\]
where \(Y_1\) is Bernoulli and \(Y_2\) is Poisson with a conditional intensity
defined below.

The output follows the same structure as the ZIP report: selection
frequencies for \(X_1,\ldots,X_{18}\), the true-positive rate, the
false-positive rate, and the mean cardinality of the selected set. We also add
FDR-SSP curves comparing `mfscreen` with three marginal screening competitors.

# Mathematical simulation framework

Let \(n=1500\), \(p=2000\), \(D=10\), and \(N=80\) Monte Carlo replications.
For observation \(i\), write
\[
X_i=(X_i^{(1)},\ldots,X_i^{(p)}).
\]

The active response-generating variables are \(X^{(1)},\ldots,X^{(10)}\).
As in the ZIP screening table, the variables counted as positives are
\[
\mathcal A = \{1,\ldots,14\},
\]
because \(X^{(11)},\ldots,X^{(14)}\) are noisy copies of active Gaussian
variables. The null set used for false-positive and FDR calculations is
\[
\mathcal A^c = \{15,\ldots,2000\}.
\]

## First five Gaussian variables

Let
\[
(X^{(1)},\ldots,X^{(5)}) \sim \mathcal N_5(0,\Sigma_\rho),
\qquad
(\Sigma_\rho)_{jk}=\rho^{|j-k|}.
\]
The executable parameter setting in the source ZIP file is
\[
\rho = 0.
\]

## Additional variables entering the response

Independently of the first five Gaussian variables, generate
\[
X^{(6)} \sim \mathrm{Bernoulli}(0.35),
\qquad
X^{(7)} \sim \chi^2_2,
\]
\[
X^{(8)}=X^{(1)}P,\quad P\sim\mathrm{Poisson}(2),
\]
\[
X^{(9)}=X^{(2)}G,\quad G\sim \mathcal N(1,1),
\]
and
\[
X^{(10)}\sim t_{14}.
\]

## Correlated non-generating variables

Let
\[
Z_3\sim\mathcal N(0,0.4^2),\quad
Z_4\sim\mathcal N(0,0.02^2),\quad
Z_5\sim\mathcal N(0,0.35^2),\quad
Z_6\sim\mathcal N(0,0.55^2),
\]
independent of the variables above. Define
\[
X^{(11)}=X^{(1)}+Z_3,\quad
X^{(12)}=X^{(3)}+Z_4,\quad
X^{(13)}=X^{(4)}+Z_5,\quad
X^{(14)}=X^{(5)}+Z_6,
\]
and
\[
X^{(15)}=Z_3,\quad X^{(16)}=Z_4,\quad
X^{(17)}=Z_5,\quad X^{(18)}=Z_6.
\]
The variables \(X^{(11)},\ldots,X^{(14)}\) are correlated with active
variables but do not enter the response equation. Variables
\(X^{(15)},\ldots,X^{(18)}\) are null variables and are counted as false
positives when selected.

## Remaining null block

The remaining variables
\[
X^{(19)},\ldots,X^{(2000)}
\]
are independent of \((Y,X^{(1)},\ldots,X^{(18)})\). They form
a stationary Gaussian sequence with covariance
\[
\operatorname{Cov}(X^{(18+j)},X^{(18+k)})=\rho_{\tilde{}}^{|j-k|},
\qquad \rho_{\tilde{}}=0.7.
\]

## Zero-inflated Poisson response

The coefficient vector is the one used in the executable part of the ZIP
file:
\[
\theta^*=(1,2,\ldots,11)^T.
\]
For the first ten covariates, define the linear predictor
\[
\eta_i=\theta_0^*+\sum_{j=1}^{10}\theta_j^*X_i^{(j)}.
\]
The zero-inflation component is
\[
Y_{1i}\mid X_i \sim
\mathrm{Bernoulli}\left(\pi_i\right),
\qquad
\pi_i=\frac{\exp(\eta_i)}{1+\exp(\eta_i)}.
\]

The Poisson component is
\[
Y_{2i}\mid X_i \sim \mathrm{Poisson}(\lambda_i),
\qquad
\lambda_i=|\eta_i|.
\]
Conditionally on \(X_i\), \(Y_{1i}\) and \(Y_{2i}\) are generated
independently. The observed response is
\[
Y_i=Y_{1i}Y_{2i}.
\]
Thus
\[
\mathbb P(Y_i=0\mid X_i)
=(1-\pi_i)+\pi_i\exp(-|\eta_i|),
\]
and for every integer \(y\geq 1\),
\[
\mathbb P(Y_i=y\mid X_i)
=\pi_i\exp(-|\eta_i|)
\frac{|\eta_i|^y}{y!}.
\]

# Screening rules and criteria

For `mfscreen`, the package returns reciprocal scores \(D_j^2\), with smaller
values indicating stronger marginal association. At nominal level \(q\),
variable \(j\) is selected when
\[
D_j^2 \leq \frac{1}{\gamma_q^2},
\qquad
\gamma_q=\Phi^{-1}(1-q/2).
\]

The ZIP-type screening table uses the source-file threshold
\[
q_n=0.15.
\]
For a selected set \(\widehat{\mathcal M}\), define
\[
\mathrm{TPR} =
\frac{1}{|\mathcal A|}\sum_{j\in\mathcal A}
\mathbf 1\{j\in\widehat{\mathcal M}\},
\]
\[
\mathrm{FPR} =
\frac{1}{|\mathcal A^c|}\sum_{j\in\mathcal A^c}
\mathbf 1\{j\in\widehat{\mathcal M}\},
\]
\[
\mathrm{FDR} =
\frac{\sum_{j\in\mathcal A^c}\mathbf 1\{j\in\widehat{\mathcal M}\}}
{\max(1,|\widehat{\mathcal M}|)},
\]
and
\[
\mathrm{SSP}=
\mathbf 1\{\mathcal A\subseteq\widehat{\mathcal M}\}.
\]

The three competitors are:

- Pearson SIS, based on marginal Pearson correlations with the observed ZIP
  response;
- Spearman SIS, based on marginal rank correlations with the observed ZIP
  response;
- Poisson-score SIS, based on the univariate Poisson score statistic for a
  count-response marginal model.

We also add Poisson-Wald SIS. It uses the same univariate Poisson marginal
model as Poisson-score SIS, but fits the slope and uses a Wald statistic.

For the FDR-SSP curves, all methods, including `mfscreen`, select the top
\(k\) variables according to their marginal ranking score. This makes the
curves directly comparable at the same selected model sizes. The calibrated
`mfscreen` threshold \(q_n=0.15\) is reported separately in the ZIP-type
screening table.

## Mathematical formulation of the competitors

For each covariate \(j\), write
\[
X_j=(X_1^{(j)},\ldots,X_n^{(j)})^T,
\qquad
Y=(Y_1,\ldots,Y_n)^T.
\]

**Pearson SIS.** The marginal Pearson statistic is
\[
R_j =
\frac{\sum_{i=1}^n (X_i^{(j)}-\bar X_j)(Y_i-\bar Y)}
{\left[\sum_{i=1}^n (X_i^{(j)}-\bar X_j)^2\right]^{1/2}
 \left[\sum_{i=1}^n (Y_i-\bar Y)^2\right]^{1/2}}.
\]
Variables are ranked by \(|R_j|\).

**Spearman SIS.** Let \(r_i^{(j)}\) be the rank of \(X_i^{(j)}\) among
\(X_1^{(j)},\ldots,X_n^{(j)}\), and let \(s_i\) be the rank of \(Y_i\) among
\(Y_1,\ldots,Y_n\). The Spearman statistic is
\[
R_j^S =
\frac{\sum_{i=1}^n (r_i^{(j)}-\bar r_j)(s_i-\bar s)}
{\left[\sum_{i=1}^n (r_i^{(j)}-\bar r_j)^2\right]^{1/2}
 \left[\sum_{i=1}^n (s_i-\bar s)^2\right]^{1/2}}.
\]
Variables are ranked by \(|R_j^S|\).

**Poisson-score SIS.** For each \(j\), consider the marginal Poisson GLM
\[
Y_i\mid X_i^{(j)}\sim\mathrm{Poisson}(\mu_{ij}),
\qquad
\log \mu_{ij}=\alpha_j+\beta_jX_i^{(j)}.
\]
The null hypothesis is \(H_{0j}:\beta_j=0\). Under the null model,
\[
\widehat\alpha_{0j}=\log \bar Y,
\qquad
\widehat\mu_{0i}=\bar Y.
\]
Using centered covariates, the score for \(\beta_j\) at \(H_{0j}\) is
\[
U_j=\sum_{i=1}^n (X_i^{(j)}-\bar X_j)(Y_i-\bar Y),
\]
and the Fisher information for \(\beta_j\) at the null is
\[
I_j=\bar Y\sum_{i=1}^n (X_i^{(j)}-\bar X_j)^2.
\]
The standardized score statistic is
\[
Z_j^{\mathrm{score}}=\frac{U_j}{\sqrt{I_j}},
\]
with two-sided p-value
\[
p_j^{\mathrm{score}}=2\Phi(-|Z_j^{\mathrm{score}}|).
\]

**Poisson-Wald SIS.** In the same marginal Poisson GLM, fit the unrestricted
model and obtain the maximum likelihood estimate
\((\widehat\alpha_j,\widehat\beta_j)\). Let
\(\widehat{\operatorname{se}}(\widehat\beta_j)\) be the standard error of
\(\widehat\beta_j\), computed from the inverse observed Fisher information.
The Wald statistic is
\[
Z_j^{\mathrm{Wald}}=
\frac{\widehat\beta_j}{\widehat{\operatorname{se}}(\widehat\beta_j)},
\]
with two-sided p-value
\[
p_j^{\mathrm{Wald}}=2\Phi(-|Z_j^{\mathrm{Wald}}|).
\]

The Poisson-score and Poisson-Wald tests use the same marginal Poisson model
and test the same null hypothesis \(H_{0j}:\beta_j=0\). They have the same
standard normal limiting null distribution. They are not exactly the same in
finite samples: the score statistic is evaluated at the restricted null model,
whereas the Wald statistic is evaluated at the fitted unrestricted model.

```{r functions}
n <- 1500L
N <- 80L
Dtotal <- 2000L
D <- 10L
Dtilde <- Dtotal - D - 8L
positive <- seq_len(14L)
null <- 15L:Dtotal

d <- 5L
rho <- 0
rho_tilde <- 0.7
thetastar <- 1:11
q_n <- 0.15

k_grid <- c(14L, 20L, 30L, 40L, 60L, 80L, 120L, 160L, 220L)

simulate_ar1_gaussian <- function(n, d, rho) {
  X <- matrix(rnorm(n * d), nrow = n, ncol = d)
  if (d >= 2L && abs(rho) > 0) {
    scale_innovation <- sqrt(1 - rho^2)
    for (j in 2:d) {
      X[, j] <- rho * X[, j - 1L] + scale_innovation * X[, j]
    }
  }
  X
}

simulate_zip_design_response <- function(n = n) {
  design_mat <- simulate_ar1_gaussian(n, d, rho)

  Z3 <- rnorm(n, mean = 0, sd = 0.4)
  Z4 <- rnorm(n, mean = 0, sd = 0.02)
  Z5 <- rnorm(n, mean = 0, sd = 0.35)
  Z6 <- rnorm(n, mean = 0, sd = 0.55)

  X_signal <- cbind(
    design_mat,
    rbinom(n, size = 1L, prob = 0.35),
    rchisq(n, df = 2),
    design_mat[, 1] * rpois(n, lambda = 2),
    design_mat[, 2] * rnorm(n, mean = 1, sd = 1),
    rt(n, df = 14)
  )

  eta <- as.vector(thetastar[1] + X_signal %*% thetastar[-1])
  prob <- plogis(eta)
  y1 <- rbinom(n, size = 1L, prob = prob)
  y2 <- rpois(n, lambda = abs(eta))
  y <- y1 * y2

  X_extra <- cbind(
    design_mat[, 1] + Z3,
    design_mat[, 3] + Z4,
    design_mat[, 4] + Z5,
    design_mat[, 5] + Z6,
    Z3, Z4, Z5, Z6,
    simulate_ar1_gaussian(n, Dtilde, rho_tilde)
  )

  X <- cbind(X_signal, X_extra)
  colnames(X) <- paste0("X", seq_len(ncol(X)))
  list(X = X, y = y, y1 = y1, y2 = y2, probability = prob, eta = eta)
}

pearson_pvalues <- function(X, y) {
  n <- nrow(X)
  yc <- y - mean(y)
  Xc <- sweep(X, 2L, colMeans(X), check.margin = FALSE)
  denom <- sqrt(colSums(Xc^2) * sum(yc^2))
  r <- colSums(Xc * yc) / denom
  r[!is.finite(r)] <- 0
  r <- pmin(pmax(r, -1 + 1e-15), 1 - 1e-15)
  stat <- abs(r) * sqrt((n - 2) / pmax(1 - r^2, .Machine$double.eps))
  p <- 2 * pt(-stat, df = n - 2)
  p[!is.finite(p)] <- 1
  p
}

spearman_pvalues <- function(X, y) {
  X_rank <- apply(X, 2L, rank, ties.method = "average")
  y_rank <- rank(y, ties.method = "average")
  pearson_pvalues(X_rank, y_rank)
}

poisson_score_pvalues <- function(X, y) {
  mu_hat <- mean(y)
  if (mu_hat <= 0) {
    return(rep(1, ncol(X)))
  }
  Xc <- sweep(X, 2L, colMeans(X), check.margin = FALSE)
  score <- colSums(Xc * (y - mu_hat))
  info <- mu_hat * colSums(Xc^2)
  z <- score / sqrt(pmax(info, .Machine$double.eps))
  p <- 2 * pnorm(-abs(z))
  p[!is.finite(p)] <- 1
  p
}

poisson_wald_pvalues <- function(X, y, max_iter = 8L, tol = 1e-7) {
  p <- ncol(X)
  alpha <- rep(log(mean(y) + 1e-8), p)
  beta <- rep(0, p)

  for (iter in seq_len(max_iter)) {
    eta <- sweep(X, 2L, beta, `*`)
    eta <- sweep(eta, 2L, alpha, `+`)
    eta <- pmin(pmax(eta, -30), 30)
    mu <- exp(eta)

    resid <- y - mu
    s0 <- colSums(resid)
    s1 <- colSums(X * resid)
    i00 <- colSums(mu)
    i01 <- colSums(X * mu)
    i11 <- colSums(X^2 * mu)
    det <- pmax(i00 * i11 - i01^2, .Machine$double.eps)

    delta_alpha <- (i11 * s0 - i01 * s1) / det
    delta_beta <- (-i01 * s0 + i00 * s1) / det
    delta_alpha <- pmin(pmax(delta_alpha, -2), 2)
    delta_beta <- pmin(pmax(delta_beta, -2), 2)

    alpha <- alpha + delta_alpha
    beta <- beta + delta_beta

    if (max(abs(delta_alpha), abs(delta_beta), na.rm = TRUE) < tol) {
      break
    }
  }

  eta <- sweep(X, 2L, beta, `*`)
  eta <- sweep(eta, 2L, alpha, `+`)
  eta <- pmin(pmax(eta, -30), 30)
  mu <- exp(eta)
  i00 <- colSums(mu)
  i01 <- colSums(X * mu)
  i11 <- colSums(X^2 * mu)
  det <- pmax(i00 * i11 - i01^2, .Machine$double.eps)
  se_beta <- sqrt(i00 / det)
  z <- beta / pmax(se_beta, .Machine$double.eps)
  pval <- 2 * pnorm(-abs(z))
  pval[!is.finite(pval)] <- 1
  pval
}

selection_metrics <- function(selected) {
  selected_count <- sum(selected)
  false_discoveries <- sum(selected[null])
  true_positives <- sum(selected[positive])
  c(
    fdr = false_discoveries / max(1L, selected_count),
    fpr = false_discoveries / length(null),
    tpr = true_positives / length(positive),
    ssp = as.numeric(all(selected[positive])),
    selected_count = selected_count
  )
}

top_k_selected <- function(pvalues, k) {
  selected <- rep(FALSE, length(pvalues))
  selected[order(pvalues, decreasing = FALSE)[seq_len(k)]] <- TRUE
  selected
}

mfscreen_selected_from_scores <- function(scores, q) {
  gamma <- qnorm(1 - q / 2)
  is.finite(scores) & scores <= 1 / gamma^2
}
```

# Monte Carlo computation

For each replication, one full design matrix and one ZIP response are
generated. The same data set is used for `mfscreen` and all competitors.

```{r monte-carlo}
run_one_replication <- function(replication) {
  dat <- simulate_zip_design_response(n)
  X <- dat$X
  y <- dat$y

  mf_scores <- screening_test_matrix(X, y, q = q_n)$scores
  mf_qn_selected <- mfscreen_selected_from_scores(mf_scores, q_n)

  method_pvalues <- list(
    mfscreen = mf_scores,
    `Pearson SIS` = pearson_pvalues(X, y),
    `Spearman SIS` = spearman_pvalues(X, y),
    `Poisson-score SIS` = poisson_score_pvalues(X, y),
    `Poisson-Wald SIS` = poisson_wald_pvalues(X, y)
  )

  curve_results <- do.call(rbind, lapply(names(method_pvalues),
    function(method) {
      do.call(rbind, lapply(k_grid, function(k) {
        selected <- top_k_selected(method_pvalues[[method]], k)
        data.frame(
          replication = replication,
          method = method,
          parameter = k,
          parameter_label = paste0("top ", k),
          t(selection_metrics(selected)),
          row.names = NULL
        )
      }))
    }))

  list(
    curves = curve_results,
    first18 = mf_qn_selected[1:18],
    qn_metrics = selection_metrics(mf_qn_selected)
  )
}

replications <- lapply(seq_len(N), run_one_replication)

raw_results <- do.call(rbind, lapply(replications, `[[`, "curves"))
first18_selected <- do.call(rbind, lapply(replications, `[[`, "first18"))
qn_metrics <- do.call(rbind, lapply(replications, `[[`, "qn_metrics"))

summary_results <- aggregate(
  cbind(fdr, fpr, tpr, ssp, selected_count) ~ method + parameter + parameter_label,
  data = raw_results,
  FUN = mean
)
summary_sd <- aggregate(
  cbind(fdr, fpr, tpr, ssp, selected_count) ~ method + parameter + parameter_label,
  data = raw_results,
  FUN = sd
)
names(summary_sd)[4:8] <- paste0(names(summary_sd)[4:8], "_sd")
summary_results <- merge(
  summary_results,
  summary_sd,
  by = c("method", "parameter", "parameter_label")
)
summary_results <- summary_results[order(summary_results$method,
                                         summary_results$parameter), ]

write.csv(raw_results, file.path(out_dir, "zip_response_fdr_ssp_raw.csv"),
          row.names = FALSE)
write.csv(summary_results, file.path(out_dir, "zip_response_fdr_ssp_summary.csv"),
          row.names = FALSE)
```

# ZIP-type screening output

The following table has the same structure as the output table in the ZIP
simulation file. It reports the selection frequency for \(X_1,\ldots,X_{18}\)
at \(q_n=0.15\), followed by the average true-positive rate, the average
false-positive rate, and the average size of the selected set.

```{r zip-output-table}
first18_names <- c(
  "X1", "X2", "X3", "X4", "X5", "X6_bern", "X7_chisq2",
  "X8=X1*P(2)", "X9=N(1,1)*X2", "X10=Student(14)",
  "X11=X1+Z3", "X12=X3+Z4", "X13=X4+Z5", "X14=X5+Z6",
  "X15=Z3", "X16=Z4", "X17=Z5", "X18=Z6"
)

screen_18 <- colMeans(first18_selected)
resultats_screening <- c(
  screen_18,
  taux_vrais_positifs = mean(qn_metrics[, "tpr"]),
  taux_faux_positifs = mean(qn_metrics[, "fpr"]),
  cardinal_moyen_ensemble_indices_selectionnes =
    mean(qn_metrics[, "selected_count"])
)

resultats_screening_table <- data.frame(
  quantity = c(first18_names,
               "taux_vrais_positifs",
               "taux_faux_positifs",
               "cardinal_moyen_ensemble_indices_selectionnes"),
  value = as.numeric(resultats_screening)
)

write.csv(resultats_screening_table,
          file.path(out_dir, "zip_response_resultats_screening_q015.csv"),
          row.names = FALSE)

knitr::kable(
  resultats_screening_table,
  digits = 3,
  caption = paste0("ZIP-design screening results: N = ", N,
                   ", n = ", n, ", q_n = 0.15.")
)
```

The expected cardinality under perfect recovery of the 14 positive variables
and nominal false-positive rate \(q_n=0.15\) is
\[
14 + 0.15\times 1986 = 311.9.
\]

```{r qn-summary}
qn_summary <- data.frame(
  metric = c("FDR", "FPR", "TPR", "SSP", "Mean selected cardinality"),
  value = c(
    mean(qn_metrics[, "fdr"]),
    mean(qn_metrics[, "fpr"]),
    mean(qn_metrics[, "tpr"]),
    mean(qn_metrics[, "ssp"]),
    mean(qn_metrics[, "selected_count"])
  )
)

write.csv(qn_summary, file.path(out_dir, "zip_response_mfscreen_q015_summary.csv"),
          row.names = FALSE)

knitr::kable(
  qn_summary,
  digits = 3,
  caption = "Global mfscreen metrics at the source threshold q_n = 0.15."
)
```

# Competitor comparison

```{r comparison-table}
knitr::kable(
  summary_results[order(summary_results$method, summary_results$parameter),
                  c("method", "parameter_label", "fdr", "ssp", "tpr",
                    "selected_count")],
  digits = 3,
  caption = paste0("FDR-SSP curve summary over N = ", N, " replications.")
)
```

# FDR-SSP curves

```{r fdr-ssp-plot}
curve_plot <- ggplot(
  summary_results,
  aes(x = fdr, y = ssp, colour = method, shape = method)
) +
  geom_path(linewidth = 0.85, alpha = 0.9) +
  geom_point(size = 2.2, alpha = 0.95) +
  scale_colour_manual(values = method_colours) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.2)) +
  labs(
    x = "Mean FDR",
    y = "SSP: probability of selecting X1,...,X14",
    colour = NULL,
    shape = NULL,
  caption = "Each method varies the same top-k selected model size grid."
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.caption = element_text(hjust = 0)
  )

curve_plot

ggsave(file.path(out_dir, "zip_response_fdr_ssp_curve.pdf"), curve_plot,
       width = 7.2, height = 4.8)
ggsave(file.path(out_dir, "zip_response_fdr_ssp_curve.png"), curve_plot,
       width = 7.2, height = 4.8, dpi = 180)
```

```{r fdr-selected-plot}
selected_plot <- ggplot(
  summary_results,
  aes(x = selected_count, y = fdr, colour = method, shape = method)
) +
  geom_path(linewidth = 0.85, alpha = 0.9) +
  geom_point(size = 2.2, alpha = 0.95) +
  scale_colour_manual(values = method_colours) +
  labs(
    x = "Mean number of selected variables",
    y = "Mean FDR",
    colour = NULL,
    shape = NULL
  ) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

selected_plot

ggsave(file.path(out_dir, "zip_response_fdr_selected_count.pdf"), selected_plot,
       width = 7.2, height = 4.8)
ggsave(file.path(out_dir, "zip_response_fdr_selected_count.png"), selected_plot,
       width = 7.2, height = 4.8, dpi = 180)
```

# Conclusion

The ZIP-response experiment uses the ZIP-design covariates without changing
the active/null structure of the screening table. At the source threshold
\(q_n=0.15\), the mfscreen false-positive rate is directly comparable with the
nominal target, while the reported selection frequencies show which of
\(X_1,\ldots,X_{18}\) are retained. The FDR-SSP curves compare all methods at
the same top-\(k\) selected model sizes, so the curves are directly comparable.
