---
title: "Efficiency of model-free screening under shared volatility"
author: "mfscreen simulation"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: pdflatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 7.2,
  fig.height = 4.8
)

library(ggplot2)

result_dir <- "shared_volatility_active_signal_results"
summary_results <- read.csv(file.path(result_dir, "summary_results.csv"))
raw_results <- read.csv(file.path(result_dir, "raw_results.csv"))

summary_results$q_label <- ifelse(
  summary_results$q %in% c(0.001, 0.01, 0.05, 0.10, 0.30),
  as.character(summary_results$q),
  ""
)

q_levels <- sort(unique(summary_results$q))
n <- 1000L
p <- 1000L
s <- 5L
n_rep <- length(unique(raw_results$replication))
```

# Motivation

The previous simulations compare screening methods in standard settings where
the null predictors are close to the assumptions behind Pearson and Spearman
correlation p-values. In those cases, all methods can have good FPR calibration.

Here the goal is to create a more diagnostic simulation for the efficiency of
the reciprocal model-free statistic. The question is:

> Can a method retain the true active variables while avoiding false positives
> caused by a common volatility factor?

This is relevant for high-dimensional screening because some variables can be
selected not because they have a mean association with the response, but because
they share observation-level volatility, outliers, or scale heterogeneity with
the response. A good screening statistic should not confuse this volatility
artifact with a useful marginal signal.

# Competing Methods

We compare the same three marginal screening rules.

## Model-free reciprocal statistic

For predictor \(X_j\), let
\[
z_{ij} = X_{ij} - \bar X_j, \qquad u_i = Y_i - \bar Y.
\]
The reciprocal statistic is
\[
D_j^2 =
\frac{\sum_i z_{ij}^2 u_i^2}{\left(\sum_i z_{ij}u_i\right)^2}
-
\frac{2\sum_i z_{ij}^3 u_i}
{\left(\sum_i z_{ij}u_i\right)\left(\sum_i z_{ij}^2\right)}
+
\frac{\sum_i z_{ij}^4}{\left(\sum_i z_{ij}^2\right)^2}.
\]
Small values indicate stronger marginal evidence. At nominal level \(q\), the
selection threshold is
\[
D_j^2 \leq \frac{1}{\{\Phi^{-1}(1-q/2)\}^2}.
\]

The denominator uses the empirical second moment of \(z_{ij}u_i\), which makes
the statistic sensitive to heteroskedasticity in a way that ordinary
correlation p-values are not.

## Pearson SIS

Pearson SIS computes the ordinary Pearson correlation \(r_j\) between \(X_j\)
and \(Y\), converts it to a two-sided p-value using the usual \(t\)-test, and
selects if
\[
p_j \leq q.
\]
This is efficient under homoskedastic linear Gaussian assumptions, but the
p-value is not designed for common volatility or strong scale heterogeneity.

## Spearman SIS

Spearman SIS applies the same p-value rule to ranks. It is less sensitive to
extreme values than Pearson SIS, but the rank p-value is still a simple
correlation calibration and can be affected by common volatility.

# Simulation Design

We use fixed
\[
n = `r n`, \qquad p = `r p`,
\]
with active set
\[
S = \{1,2,3,4,5\}.
\]
The number of Monte Carlo replications is
\[
B = `r n_rep`.
\]
The q grid is
\[
`r paste(q_levels, collapse = ", ")`.
\]

For the active variables,
\[
X_1,\ldots,X_5 \sim N(0,1),
\]
and the mean signal is
\[
\mu(X) =
0.30 X_1 - 0.27 X_2 + 0.24 X_3 - 0.21 X_4 + 0.18 X_5.
\]

We compare two scenarios.

## Clean homoskedastic scenario

All inactive predictors are independent standard normal variables and
\[
Y = \mu(X) + \varepsilon, \qquad \varepsilon \sim N(0,1).
\]
This is a favorable baseline for Pearson SIS.

## Shared-volatility scenario

Let
\[
A_i = \exp(0.5 Z_i), \qquad Z_i \sim N(0,1).
\]
The active variables remain unscaled standard normal variables. The inactive
variables and the noise component of the response share the same volatility:
\[
X_{ij} = A_i E_{ij}, \quad j \notin S,
\]
and
\[
Y_i = \mu(X_i) + A_i \varepsilon_i.
\]
Here \(E_{ij}\), \(Z_i\), and \(\varepsilon_i\) are independent standard normal
variables.

Conditionally on the volatility factor, inactive predictors have no mean
association with the response. However, the shared scale factor can generate
large products \(X_{ij}Y_i\) and can make ordinary correlation p-values
anti-conservative.

# How Points Are Drawn

For a fixed method and q level, each Monte Carlo replication produces a
selected set \(\widehat S_q\). We compute
\[
\mathrm{FPR}_b(q) =
\frac{|\widehat S_q^{(b)} \cap S^c|}{|S^c|}
\]
and
\[
\mathrm{SSP}_b(q) =
1\{S \subseteq \widehat S_q^{(b)}\}.
\]
The plotted point is
\[
\left(
\frac{1}{B}\sum_b \mathrm{FPR}_b(q),
\frac{1}{B}\sum_b \mathrm{SSP}_b(q)
\right).
\]

The same selected set also gives a model-size measure:
\[
|\widehat S_q|.
\]

# Results

## Efficiency frontier

```{r frontier}
ggplot(
  summary_results,
  aes(x = fpr, y = ssp, colour = method, shape = method)
) +
  geom_path(aes(group = method), linewidth = 0.7) +
  geom_point(size = 2.2) +
  geom_text(
    aes(label = q_label),
    size = 2.5,
    nudge_y = 0.025,
    show.legend = FALSE
  ) +
  facet_wrap(~ scenario) +
  coord_cartesian(xlim = c(0, max(summary_results$fpr) * 1.05),
                  ylim = c(0, 1.04)) +
  labs(
    x = "False positive rate",
    y = "Sure screening property",
    colour = "Method",
    shape = "Method",
    title = "Efficiency frontier under shared volatility",
    subtitle = paste0("n = ", n, ", p = ", p,
                      ", active variables = ", s,
                      ", replications = ", n_rep)
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

In the clean homoskedastic scenario, all three methods are similar. This is
expected: the competitors are being used in a setting close to their ideal
calibration.

The shared-volatility scenario is more revealing. Pearson SIS reaches high SSP,
but does so with many false positives. For example, at \(q=0.05\), Pearson SIS
has SSP about 0.98 but FPR about 0.20. In contrast, `mfscreen` has the same SSP
about 0.98 but FPR about 0.056. Spearman SIS is more robust than Pearson, but
still has FPR about 0.069 at \(q=0.05\), larger than `mfscreen`.

This is the efficiency advantage: for comparable sure screening probability,
`mfscreen` produces a much smaller candidate set under shared volatility.

## FPR calibration

```{r fpr-q}
ggplot(
  summary_results,
  aes(x = q, y = fpr, colour = method, shape = method)
) +
  geom_line(
    data = data.frame(q = q_levels, fpr = q_levels),
    aes(x = q, y = fpr),
    inherit.aes = FALSE,
    linetype = "dashed",
    colour = "grey45"
  ) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ scenario) +
  scale_x_log10(breaks = q_levels) +
  labs(
    x = "Nominal q level",
    y = "False positive rate",
    colour = "Method",
    shape = "Method",
    title = "FPR control by q"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

The dashed line is the ideal calibration \(\mathrm{FPR}=q\). In the clean
scenario, all methods follow this line closely. Under shared volatility,
`mfscreen` remains close to the nominal q level, while Pearson SIS is strongly
anti-conservative and Spearman SIS is moderately anti-conservative.

At \(q=0.01\), Pearson SIS has FPR about 0.094, roughly nine times the nominal
level. At the same q level, `mfscreen` has FPR about 0.012. This difference is
not a minor plotting artifact: it means Pearson includes many inactive
variables before the downstream model-building step.

## Number of selected variables

```{r selected-count}
ggplot(
  summary_results,
  aes(x = q, y = selected_count, colour = method, shape = method)
) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ scenario) +
  scale_x_log10(breaks = q_levels) +
  labs(
    x = "Nominal q level",
    y = "Average selected variables",
    colour = "Method",
    shape = "Method",
    title = "Model size induced by screening"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position = "bottom")
```

The model-size plot gives a practical interpretation of FPR inflation. In the
shared-volatility scenario at \(q=0.05\), `mfscreen` selects about 61 variables
on average, Spearman SIS selects about 73, and Pearson SIS selects about 204.
All three methods have high SSP there, but Pearson pays for it by passing a
much larger candidate set to later analysis.

# Numerical Summary

```{r table}
summary_table <- summary_results[order(summary_results$scenario,
                                       summary_results$q,
                                       summary_results$method),
                                 c("scenario", "q", "method", "fpr", "ssp",
                                   "selected_count")]
knitr::kable(
  summary_table,
  digits = 4,
  row.names = FALSE,
  caption = "Mean FPR, SSP, and selected model size over Monte Carlo replications."
)
```

# Conclusion

This simulation isolates a situation where ordinary correlation screening is
not just less robust but practically inefficient. Shared volatility creates
many false positives for Pearson SIS and, to a lesser extent, Spearman SIS.
The model-free reciprocal statistic keeps the false positive rate close to the
target while retaining the active variables with similar probability. This
supports using `mfscreen` as a first-stage screen when high-dimensional data may
contain scale heterogeneity, common volatility, or outlier-prone observations.
