---
title: "Gaussian null study: convergence rate of the false-positive probability"
author: "J. Dedecker, M.-L. Taupin, V. Runge and A.-S. Tocquet"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
  html_document:
    toc: true
    toc_depth: 2
---

# Objective

This study considers the simplest null setting:

\[
X_i \sim \mathcal N(0,1), \qquad
Y_i \sim \mathcal N(0,1), \qquad
X_i \perp Y_i.
\]

Each simulated data set contains one Gaussian predictor column \(X\) and one
independent Gaussian response \(Y\). Repeating the experiment over many
independent data sets is equivalent to studying many independent null columns.

For a sample size \(n\), define the finite-sample false-positive probability

\[
\operatorname{FPR}_n =
\mathbb P\left\{
  \left|\widehat\gamma_n\right|
  \geq \Phi^{-1}(1-q/2)
\right\}.
\]

The quantity expected to decrease at rate \(C/\sqrt{n}\) is the finite-sample
bias relative to the nominal level:

\[
\left|\operatorname{FPR}_n-q\right|.
\]

Therefore, the target slope in the log-log regression of
\(\left|\widehat{\operatorname{FPR}}_n-q\right|\) against \(n\) is \(-1/2\).

# Important Monte Carlo point

With a fixed number of simulations \(B\), the Monte Carlo error of
\(\widehat{\operatorname{FPR}}_n\) is of order \(1/\sqrt{B}\), independently
of \(n\). It will eventually mask a true \(1/\sqrt{n}\) bias.

For this reason, this document uses a number of simulated null columns
\(B_n\) proportional to \(n\). The resulting Monte Carlo standard error is
also proportional to \(1/\sqrt n\), but is chosen to be a small fraction of
the target rate.

# Parameters

```{r parameters}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  cache.path = "gaussian_null_fpr_rate_cache/"
)

library(parallel)
library(knitr)

# Nominal two-sided false-positive rate.
q <- 0.50

# Sample sizes for the log-log rate study.
n_values <- round(1.5^(2:9) * 200)

# The desired Monte Carlo standard error is approximately
# mc_error_fraction / sqrt(n).
mc_error_fraction <- 0.15

# For a Bernoulli(q) selection indicator,
# sqrt(q * (1 - q) / B_n) <= mc_error_fraction / sqrt(n)
# is obtained with B_n >= q(1-q)n / mc_error_fraction^2.
B_by_n <- ceiling(q * (1 - q) * n_values / mc_error_fraction^2)

# For a quick test, use the following instead:
# B_by_n <- rep(2000L, length(n_values))

base_seed <- 20260701L
n_cores <- min(8L, parallel::detectCores(logical = FALSE))

data.frame(n = n_values, B_n = B_by_n)
```

# Original model-free screening test

```{r screening-function}
screening_test_ML <- function(X, Y, q_n = 0.10) {
  X <- as.matrix(X)
  Y <- as.numeric(Y)

  n <- nrow(X)

  if (ncol(X) != 1L) {
    stop("This Gaussian-null study expects exactly one predictor column.")
  }
  if (length(Y) != n) {
    stop("Y must have the same number of observations as rows of X.")
  }

  x_mean <- mean(X[, 1L])
  y_mean <- mean(Y)

  x_centered <- X[, 1L] - x_mean
  y_centered <- Y - y_mean

  sxx <- sum(x_centered^2)
  if (sxx <= 0) {
    stop("X has zero sample variance.")
  }

  tau_hat <- sum(x_centered * y_centered) / sxx
  b_hat <- y_mean - tau_hat * x_mean
  residuals_hat <- Y - b_hat - tau_hat * X[, 1L]

  var_x_n <- sxx / n
  V_hat <- mean(x_centered^2 * residuals_hat^2) / var_x_n^2

  gamma_hat <- sqrt(n) * tau_hat / sqrt(V_hat)
  threshold_gamma <- qnorm(1 - q_n / 2)

  list(
    score = gamma_hat,
    selected = abs(gamma_hat) >= threshold_gamma,
    threshold = threshold_gamma
  )
}
```

# One Gaussian-null data set

```{r one-simulation}
run_one_null_simulation <- function(seed, n, q) {
  set.seed(seed)

  X <- rnorm(n)
  Y <- rnorm(n)

  screening_test_ML(
    X = matrix(X, ncol = 1L),
    Y = Y,
    q_n = q
  )$selected
}
```

# Estimate the FPR for one sample size

Each seed is explicit. The result is reproducible regardless of the number
of processor cores and the scheduling of parallel tasks.

```{r fpr-for-one-n}
estimate_fpr_at_n <- function(n, B, q, seed_offset, mc.cores = 1L) {
  seeds <- as.integer(seed_offset + seq_len(B))

  one_draw <- function(seed) {
    run_one_null_simulation(seed = seed, n = n, q = q)
  }

  selected <- if (.Platform$OS.type == "windows" || mc.cores <= 1L) {
    unlist(lapply(seeds, one_draw), use.names = FALSE)
  } else {
    unlist(
      parallel::mclapply(
        X = seeds,
        FUN = one_draw,
        mc.cores = mc.cores,
        mc.set.seed = FALSE,
        mc.preschedule = FALSE
      ),
      use.names = FALSE
    )
  }

  fpr_hat <- mean(selected)

  list(
    n = n,
    B = B,
    FPR_hat = fpr_hat,
    MC_SE = sqrt(fpr_hat * (1 - fpr_hat) / B)
  )
}
```

# Monte Carlo study

A different seed block is allocated to each sample size. This avoids overlap
between the random-number streams used at different \(n\).

```{r monte-carlo, cache=TRUE}
seed_offsets <- cumsum(c(base_seed, head(B_by_n, -1L) + 1000L))

fpr_results <- Map(
  f = function(n, B, seed_offset) {
    estimate_fpr_at_n(
      n = n,
      B = B,
      q = q,
      seed_offset = seed_offset,
      mc.cores = n_cores
    )
  },
  n = n_values,
  B = B_by_n,
  seed_offset = seed_offsets
)

fpr_rate_summary <- do.call(
  rbind,
  lapply(fpr_results, as.data.frame)
)

fpr_rate_summary$FPR_minus_q <- fpr_rate_summary$FPR_hat - q
fpr_rate_summary$abs_FPR_minus_q <- abs(fpr_rate_summary$FPR_minus_q)
fpr_rate_summary$sqrt_n_times_abs_FPR_minus_q <-
  sqrt(fpr_rate_summary$n) * fpr_rate_summary$abs_FPR_minus_q
fpr_rate_summary$signal_to_MC_noise <-
  fpr_rate_summary$abs_FPR_minus_q / fpr_rate_summary$MC_SE
```

# FPR convergence-rate table

```{r fpr-summary}
knitr::kable(
  fpr_rate_summary,
  digits = 6,
  caption = paste0(
    "Gaussian-null FPR study: q = ", q,
    "; the number of null columns B_n increases with n"
  )
)
```

# Log-log rate estimate

The regression uses points whose estimated FPR deviation is at least twice
the corresponding Monte Carlo standard error. This prevents Monte Carlo
noise from dominating the fitted slope.

```{r rate-estimate}
usable <- with(
  fpr_rate_summary,
  abs_FPR_minus_q > 2 * MC_SE
)

if (sum(usable) >= 3L) {
  rate_fit <- lm(
    log(abs_FPR_minus_q) ~ log(n),
    data = fpr_rate_summary,
    subset = usable
  )

  rate_fit_summary <- data.frame(
    n_points_used = sum(usable),
    estimated_log_log_slope = unname(coef(rate_fit)[2L]),
    target_slope = -0.5
  )

  knitr::kable(
    rate_fit_summary,
    digits = 4,
    caption = "Estimated log-log slope of the FPR deviation"
  )
} else {
  rate_fit <- NULL
  message(
    "Too few usable points. Increase B_by_n or reduce mc_error_fraction."
  )
}
```

# Log-log plot

The dashed reference line has slope \(-1/2\). It is anchored at the first
usable point.

```{r log-log-plot, fig.width=7, fig.height=5}
positive <- fpr_rate_summary$abs_FPR_minus_q > 0

plot(
  fpr_rate_summary$n[positive],
  fpr_rate_summary$abs_FPR_minus_q[positive],
  log = "xy",
  type = "b",
  xlab = "Sample size n",
  ylab = expression(abs(hat(FPR)[n] - q)),
  main = "Gaussian-null FPR convergence"
)

if (any(usable)) {
  anchor <- which(usable)[1L]

  reference_n <- range(fpr_rate_summary$n)
  reference_error <-
    fpr_rate_summary$abs_FPR_minus_q[anchor] *
    (reference_n / fpr_rate_summary$n[anchor])^(-0.5)

  lines(reference_n, reference_error, lty = 2, lwd = 2)

  legend(
    "topright",
    legend = c("Observed |FPR_hat - q|", "Reference slope -1/2"),
    lty = c(1, 2),
    pch = c(1, NA),
    bty = "n"
  )
}
```

# Interpretation

- `FPR_hat` should be close to the nominal level `q`.
- A \(C/\sqrt n\) bias suggests that
  `sqrt_n_times_abs_FPR_minus_q` remains approximately stable.
- The fitted log-log slope should be near \(-0.5\) only over points where
  the FPR deviation is larger than Monte Carlo uncertainty.
- The value `q = 0.50` is used here because it gives a more visible rejection
  probability; the same code works for `q = 0.10`.

# Save results

```{r save-results}
saveRDS(
  list(
    parameters = list(
      q = q,
      n_values = n_values,
      B_by_n = B_by_n,
      mc_error_fraction = mc_error_fraction
    ),
    fpr_rate_summary = fpr_rate_summary
  ),
  file = "gaussian_null_fpr_convergence_results.rds"
)

write.csv(
  fpr_rate_summary,
  file = "gaussian_null_fpr_convergence_summary.csv",
  row.names = FALSE
)
```
